Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-08-18 draft.tex
Wrote 'Applicability of Method' section

%\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
% %%
%
% %%
% Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatter plot.
% %%
% By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
%%
\subsubsection{Dummy Variables}
Dummy variables abstract a categorical variable with multiple manifestations into several binary variables.
%%
Each binary variables encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables (\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data parameter.
%%\com{ToDo: Dummy-Variables}
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail (Fig.~\ref{fig:decision-tree} b).
%%
This allows for sequentially analyzing the classifications.
%%------------
\subsubsection{Implementation}
All analysis are carried out using \texttt{R}, a popular programming language for statistical calculations and visualizations.
%%
The interactive visualizations are realized using the \texttt{ggvis}\footnote{GGVIS Footnote} package.
%%
As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization parameters using UI controls, such as sliders.
%%
We implemented our application as web based prototype using \texttt{R Shiny}\footnote{R Shiny footnote}.
%%
The package combines the power of \texttt{R} with advantages of web-based applications.
%%
It allows us to quickly exchange results with our collaborating epidemiological partners.
%%
They can try out the technique without installing anything.
%%
Exchanging software becomes as easy as exchanging a hyperlink.
%%---------
\subsection{Results}
We ran the analysis using different subject groups:
\begin{enumerate}
	%%
	\item All subjects,
	%%
	\item subdivision into \emph{male} and \emph{female},
	%%
	\item subdivision into \emph{age quantiles}, yielding the groups (21, 43.8] (43.8, 54] (54, 64] (64, 90],
	%%
	\item subdivision into \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
	%%
\end{enumerate}
We plotted each group twice.
%%
The first plot shows all original variables.
%%
The second plot shows all categorical variables transformed into dichotomous dummy variables.

@@

Another observation during the analysis are very large decision trees achieving low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\subsection{Proposed Features}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/placeholder}
  \caption{Proposed features. (a) dented vertebraes are a sign of heavy stress and bone absorption. (b) Spine canal thickness is associated with herniated disks. (c) the spine canal shape can be used to characterize scoliosis and lordosis as well as the \emph{Cobb-Angles} \com{TODO CITE Cobb Angle}. }
  \label{fig:proposedFeatures}
\end{figure}
%%
\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Zitat Frau Hegenscheid fragen}.
%%
We have to capture earlier signs of pathological change in order to derive better predictive models.
%%

A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
%%
They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~(a)).
%%
This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
%%
Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~(b).
%%
Low spine canal thickness can be an indicator for an impending herniated disk.
%%
Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
%%
The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~(c).
%%

\subsection{Applicability of the Method}

Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
\com{Vorgestellte Parameter nicht ausreichend.}
\com{Weitere Parameter wichtig - z.B. Dicke der Wirbelkoerper, Durchmesser des Spinalkanals, Form der gesamten Wirbelsaeule, Cobb-Winkel}

>>>>

%\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
% %%
%
% %%
% Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatter plot.
% %%
% By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
%%
\subsubsection{Dummy Variables}
Dummy variables abstract a categorical variable with multiple manifestations into several binary variables.
%%
Each binary variables encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables (\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data parameter.
%%\com{ToDo: Dummy-Variables}
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail (Fig.~\ref{fig:decision-tree} b).
%%
This allows for sequentially analyzing the classifications.
%%------------
\subsubsection{Implementation}
All analysis are carried out using \texttt{R}, a popular programming language for statistical calculations and visualizations.
%%
The interactive visualizations are realized using the \texttt{ggvis}\footnote{GGVIS Footnote} package.
%%
As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization parameters using UI controls, such as sliders.
%%
We implemented our application as web based prototype using \texttt{R Shiny}\footnote{R Shiny footnote}.
%%
The package combines the power of \texttt{R} with advantages of web-based applications.
%%
It allows us to quickly exchange results with our collaborating epidemiological partners.
%%
They can try out the technique without installing anything.
%%
Exchanging software becomes as easy as exchanging a hyperlink.
%%
\com{TODO: Auf Eignung von RMarkdown als Report eingehen - ergebnisse koennen schnell publiziert und mit neuen Daten durchgetestet werden.}
%%---------
\subsection{Results}
We ran the analysis using different subject groups:
\begin{enumerate}
	%%
	\item All subjects,
	%%
	\item subdivision into \emph{male} and \emph{female},
	%%
	\item subdivision into \emph{age quantiles}, yielding the groups (21, 43.8] (43.8, 54] (54, 64] (64, 90],
	%%
	\item subdivision into \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
	%%
\end{enumerate}
We plotted each group twice.
%%
The first plot shows all original variables.
%%
The second plot shows all categorical variables transformed into dichotomous dummy variables.

@@

Another observation during the analysis are very large decision trees achieving low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\subsection{Proposed Features}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/placeholder}
  \caption{Proposed features. (a) dented vertebraes are a sign of heavy stress and bone absorption. (b) Spine canal thickness is associated with herniated disks. (c) the spine canal shape can be used to characterize scoliosis and lordosis as well as the \emph{Cobb-Angles} \cite{Cobb}. }
  \label{fig:proposedFeatures}
\end{figure}
%%
\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Zitat Frau Hegenscheid fragen}.
%%
We have to capture earlier signs of pathological change in order to derive better predictive models.
%%

A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
%%
They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~(a)).
%%
This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
%%
Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~(b).
%%
Low spine canal thickness can be an indicator for an impending herniated disk.
%%
Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
%%
The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~(c).
%% ------------------------------------
\subsection{Applicability of the Method}
%%
Classification methods based on decision trees showed to be useful to assess the discriminative power of a variable set.
%%
Their ability to assess variable combination makes them more powerful than correlation coefficients calculated for each variable.
%%
This advantage comes with a much more complex output, making their results harder to assess and to abstract.
%%
Our method to plot derived metrics and custom tailored error measures proved to be effective.
%%
Huge result spaces could be navigated fast using our brushable visualization.
%%
Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
%%

As described in a prior work \cite{Klemm2014}, the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
